RL-Musician: A Tool for Music Composition with Reinforcement Learning

### 2019-10-16 docs/paper/paper.tex
docs: Revise draft of paper

\title{RL-Musician: A Tool for Music Composition with Deep Reinforcement Learning}

@@

\begin{abstract}
The notion of creativity is wider than generating something inspired by reference pieces of art. It also includes generation of something that meets criteria of being an art, but is not backed by existing pieces. On the one hand, supervised machine learning is limited by the former scope, because it requires a dataset. Reinforcement learning can be applied instead if the goal is to reveal more aspects of creativity. On the other hand, numerous recent advances in music generation are achieved with supervised deep learning. In this paper, a method that uses both reinforcement learning and deep learning is developed for automated music composition. An agent with neural network as actor model is trained to compose musical pieces by interacting with a piano roll environment. The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory. No dataset is involved, because cross-entropy method is used instead of gradient-based methods.
\end{abstract}

@@

Brief outline of the approach is as follows. There is an environment with 2D table representing piano roll which means that rows of the table correspond to notes and columns of the table correspond to time steps. An agent observes current time step and some previous time steps. An action of the agent is playing a note at the current time step. After certain number of actions are done, a movement to the next time step happens. When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions. These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).

@@

In algorithmic composition domain, numerous links between reinforcement learning and deep learning are established.

Reinforcement learning can be used for altering weights of recurrent neural networks trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}. The goal is to make generated pieces more structured and conformed with music theory rules. To define an environment, let its state be composed of recurrent neural networks states and previously played notes, let an action be an output for current time step and let reward depend on both evaluational rules and probability of output according to initial RNN. Rewards are granted immediately after a step and so DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred over cross-entropy method as a training algorithm.

Usually, softmax activation function is used in the last layer of generative RNN. However, it is possible to consider a family of activation functions parametrized by one parameter $t$ such that output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$). Also, input vector can be extended by introducing an additional part indicating origination of initial input (so called plan). Then an environment can be defined so that state is a pair of $t$ and plan, actions are changes in either $t$ or plan, and reward depends on evaluation of produced with these settings piece \cite{kumar2019polyphonic}.

@@

Nevertheless, not all methods for automated music generation use deep learning. Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} or Markov chains \cite{farbood2001analysis}. Comprehensive review that covers plenty of approaches beyond deep learning can be found in \cite{fernandez2013ai}.

>>>>

\title{RL-Musician: A Tool for Music Composition with Reinforcement Learning}

@@

\begin{abstract}
The notion of creativity is wider than generating something inspired by reference pieces of art. It also includes generation of something that meets criteria of being an art, but is not backed by existing pieces. On the one hand, supervised machine learning is limited by the former scope, because it requires a dataset. Reinforcement learning can be applied instead if the goal is to reveal more aspects of creativity. On the other hand, numerous recent advances in music generation are achieved with supervisedly trained neural networks. In this paper, a method that involves both reinforcement learning and neural networks is developed for automated music composition. An agent with shallow neural network as actor model is trained to compose musical pieces by interacting with a piano roll environment. The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory. Cross-entropy method is used for training instead of gradient-based methods and so no datasets are needed.
\end{abstract}

@@

Brief outline of the approach is as follows. There is an environment with 2D table representing piano roll which means that rows of the table correspond to notes and columns of the table correspond to time steps. A cell of the table contains 1 if the note is sounding at this step and 0 otherwise. An agent observes vector in which for every row there is an exponential moving average over all time steps up to the current time step (inclusively). An action of the agent is playing a note at the current time step. After certain number of actions are done, a movement to the next time step happens. When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions. These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).

@@

Usage of computers for automatic composition of music dates back to 1950s and, what is more, approaches that can be labeled as algorithmic are known since Guido of Arezzo (11th century). Therefore, it is impossible to list here all studies about the subject. An interested reader can find more information in specialized reviews such as \cite{fernandez2013ai}. What is discussed below is just some examples related to reinforcement learning.

Reinforcement learning can be applied to algorithmic composition either as a standalone paradigm or in connection with deep learning.

As for the first case, there are not \cite{yi2007automatic}.

The second group (works where connections with deep learning are established) recently gained attention and several interesting approaches were developed there.

Reinforcement learning can be used for altering weights of recurrent neural networks trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}. The goal is to make generated pieces more structured and conformed with music theory rules. To define an environment, let its state be composed of recurrent neural networks states and previously played notes, let an action be an output for current time step and let reward depend on both evaluational rules and probability of output according to initial RNN. Rewards are granted immediately after a step and so DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred as a training algorithm.

Usually, softmax activation function is used in the last layer of generative RNN. However, it is possible to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted as $t$. Depending on temperature, output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$). In \cite{kumar2019polyphonic}, Boltzmann softmax is used and also input vector is extended by introducing an additional part indicating origination of initial input (so called plan). An environment is defined so that state is a pair of temperature and plan, actions are changes in either temperature or plan, and reward depends on evaluation of produced with these settings piece.

@@

% Nevertheless, not all methods for automated music generation use deep learning. Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} or HMMs (Hidden Markov Models) \cite{farbood2001analysis}
